---
title: "ReadMe"
author: "K. Huisamen"
date: "02 June 2016"
output: pdf_document
---

This document note exploratory journal entries

```{r}
# read the data
row.names(HarmfulEvents) <- HarmfulEvents$EVTYPE

```

#Read all the records?
The REFNUM show that the max is 902300, my dataset show 902297, which means I'm missing 3 records.


#Encode the events
```{r Event Types}
Event.Types <- unique(dat$EVTYPE)
```

Officially there is only 48 event types but in the database there are 898 unique event types.  To tidy the data it will encoded according to the event types as per the circular dated August 2007.

```{r Tidy Event Types}
library(dplyr)
# 48 Standard Events
ET <- list("Astronomical Low Tide" = "Astronomical Low Tide",
        "Avalanche" = "Avalanche|Avalance",
        "Blizzard" = "Blizzard",
        "Coastal Flood" = "Coastal Flood|Tidal Flood",
        "Cold/Wind Chill"= "Cold/Wind Chill|Cold",
        "Debris Flow" = "Debris Flow",
        "Dense Fog" = "Dense Fog|Fog",
        "Dense Smoke" = "Dense Smoke",
        "Drought" = "Drought",
        "Dust Devil" = "Dust Devil",
        "Dust Storm" = "Dust Storm",
        "Excessive Heat" = "Excessive Heat|Extreme Heat|Heat Wave|Record Heat",
        "Extreme Cold/Wind Chill" = "Extreme Cold/Wind Chill|Cold|Wind Chill",
        "Flash Flood" = "Flash Flood|Flash",
        "Flood" = "^Flood|Flooding|River Flood|Erosion",
        "Freezing Fog" = "Freezing Fog",
        "Frost/Freeze" = "Frost|Freez",
        "Funnel Cloud" = "Funnel Cloud",
        "Hail" = "Hail",
        "Heat" = "^Heat",
        "Heavy Rain" = "Heavy Rain",
        "Heavy Snow" = "Heavy Snow|Snow",
        "High Surf" = "High Surf|Surf",
        "High Wind" = "High Wind",
        "Hurricane/Typhoon" = "Hurricane|Typhoon",
        "Ice Storm" = "Ice Storm",
        "Lakeshore Flood" = "Lakeshore Flood",
        "Lake-Effect Snow" = "Lake-Effect Snow",
        "Lightning" = "Lightning",
        "Marine Hail" = "Marine Hail",
        "Marine High Wind" = "Marine High Wind|Heavy Seas",
        "Marine Strong Wind" = "Marine High Wind",
        "Marine Thunderstorm Wind" = "Marine Thunderstorm Wind|TSTM|MARINE MISHAP|Coastalstorm|Coastal Storm",
        "Rip Current" = "Rip Current",
        "Seiche" = "Seiche",
        "Sleet" = "Sleet",
        "Storm Tide" = "Storm Tide|Storm Surge",
        "Strong Wind" = "Strong Wind",
        "Thunderstorm Wind" = "^Thunder",
        "Tornado" = "Tornado",
        "Tropical Depression" = "Tropical Depression",
        "Tropical Storm" = "Tropical Storm",
        "Tsunami" = "Tsunami",
        "Volcanic Ash" = "Volcanic Ash",
        "Waterspout" = "Waterspout",
        "Wildfire" = "Wildfire|Forest",
        "Winter Storm" = "Winter Storm",
        "Winter Weather" = "Winter Weather")

# Tidy the events by setting all to uppercase, also to make the grepl easier.
dat <- dat %>%
        mutate(EVTYPE = toupper(EVTYPE))

# Look in the list and for...each ET load the variations, 
# match these variations in dat, append it to dat2.

dat2 <- data.frame(Event= NULL, FATALITIES = NULL, INJURIES = NULL)
for (i in 1:length(ET)){
        searchstring <- toupper(ET[[i]])
        print(searchstring)
        tidyeventlist <- dat %>%
                select(EVTYPE, FATALITIES, INJURIES) %>%
                filter(grepl(searchstring, EVTYPE)) %>%
                mutate(Event = names(ET[i]))
        dat2 <- rbind(dat2, tidyeventlist)
}

```
Assuming that if it's there's three event types for one observaton then there was three event types during that observation, so three event types would be accounted for, this will increase the data set with two additional records. The new data set is now 902342 observations, from 902297 observations originally.

# Create a summary of the events 
```{r}
# Sum the events
HarmfulEvents <- dat2 %>%
        select(Event, FATALITIES, INJURIES) %>%
        group_by(Event) %>%
        summarise(Fatalities = sum(FATALITIES), Injuries = sum(INJURIES)) %>%
        as.data.frame %>%
        mutate(Risk = (FatalityRisk*Fatalities)+(InjuryRisk*Injuries)) %>%
        filter(Risk > 0) %>%
        arrange(desc(Risk)) %>%
        mutate(Category = ntile(Risk, 5)) #Split into 5 categories

row.names(HarmfulEvents) <- HarmfulEvents$Event
HarmfulEvents <- merge(x = HarmfulEvents, y = RiskLevels, sort = FALSE)
```


#Fatalities and Injuries
To determine the relationship make a plot.

```{r Explore Fatalities and Injurues}
library(dplyr)

fi <- HarmfulEvents %>% select(Fatalities, Injuries)
plot(fi)
```

One extreme outlier, which one is it?

```{r identify outlier}
HarmfulEvents[HarmfulEvents$Fatalities >= 5000,]

```
The above was an extreme incident.

Looking into the distribution of the rest on a logscale

```{r}
plot(log10(fi))
```
As the injuries increase so does the fatalities, it also looks like there are groups of data. It might be possible to group them to present a health risk decision tree

#Histogram/Barplot as possible means to simply show ranking
```{r}
barplot(names.arg = HarmfulEvents$Event, height = log10(HarmfulEvents$Risk), horiz = TRUE, axes = FALSE, col = HarmfulEvents$Label, space = 1, main = "Event Risk Levels", ylab = "Event", xlab = "Risk", axis.lty = 1, offset = 5000)
```

#Explore using the mean values rather than the sum

```{r Explore mean}
HarmfulEvents <- dat2 %>%
        select(Event, FATALITIES, INJURIES) %>%
        group_by(Event) %>%
        summarise(Fatalities = mean(FATALITIES), Injuries = mean(INJURIES)) %>%
        as.data.frame %>%
        mutate(Risk = (FatalityRisk*Fatalities)+(InjuryRisk*Injuries)) %>%
        filter(Risk > 0) %>%
        arrange(desc(Risk)) %>%
        mutate(Category = ntile(Risk, 5)) #Split into 5 categories

row.names(HarmfulEvents) <- HarmfulEvents$Event
HarmfulEvents <- merge(x = HarmfulEvents, y = RiskLevels, sort = FALSE)

old.par <- par(no.readonly=T) 
par(las=2) #Set the labels horizontal
par(mar = c(1,15,1,1)) #Give the labels some space
barplot(names.arg = rev(HarmfulEvents$Event),
        height = rev(HarmfulEvents$Risk),
        horiz = TRUE, #Change to horizontal histogram
        axes = FALSE,
        col = HarmfulEvents$Label,
        space = 1,
        main = "Event Risk Levels",
        xlab = "Risk",
        axis.lty = 1
        )
```

row.names(HarmfulEvents) <- HarmfulEvents$Event
HarmfulEvents <- merge(x = HarmfulEvents, y = RiskLevels, sort = FALSE)

#Other possible layouts

##Hclustering and Denbdrogram, Heatmap
Explore ways to show a decision tree

```{r clustering}
fi <- HarmfulEvents %>% 
        # filter(Event != "Tornado") %>%
        select(Fatalities, Injuries)

hClustering <- fi %>%
        dist %>% hclust
plot(hClustering, hang = -1, cex = 0.6, ylab = "Height")

# Plot is way to deep, find ways to disply it
dendro <- as.dendrogram(hClustering)

nodePar <- list(lab.cex = 0.6, pch = c(NA, 19), 
                cex = 0.7, col = "blue")
plot(dendro, horiz = TRUE, nodePar = nodePar, rect.hclust(dendro, k = 2))
# Draw rectangles around clusters
rect.hclust(hc,k=2)

cutDendro <- cut(dendro, h = (hClustering$height[10]+ 1))
plot(cutDendro$lower[[33]], yaxt = "n", main = "Fatalities and Injuries")

# or, zoom in
plot(hClustering, xlim = c(1, 5), ylim = c(92000))

```

Try ggdendro
```{r}
library(ggplot2)
library(ggdendro)

ddata <- dendro_data(dendro, type = "rectangle")
p <- ggplot(segment(ddata)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) + 
  coord_flip() + 
  scale_y_reverse(expand = c(0.2, 0))
p
p + 
  coord_flip() + 
  theme_dendro()
```

##The heatmap

```{r}

dataMatrix <- as.matrix(fi[,1:2])
heatmap(dataMatrix)

# Reduce and focus on top 10 RiskLevels
dataMatrix <- as.matrix(head(fi, 40))
heatmap(dataMatrix, margins = c(10, 2))
```



##Kmeans clustering , to include more variables
```{r kmeans}
fi <- HarmfulEvents %>% select(Fatalities, Injuries)
dataMatrix <- as.matrix(fi)
kmeansObj <- kmeans(dataMatrix, centers = 4)
par(mfrow = c(1, 2))
image(t(dataMatrix)[, nrow(dataMatrix):1], yaxt = "n", main = "Original Data")
image(t(dataMatrix)[, order(kmeansObj$cluster)], yaxt = "n", main = "Clustered Data")
```

This looks like a dead end.


#Question 2 - Economic impact
Fields used for capturing is property and crop.

*  Property Damage (PROPDMG)
*  Property Damage Magnitude (PROPDMGEXP)
*  Crop Damage (CROPDMG)
*  Crop Damage Magnitude (CROPDMGEXP)

Looking at the data in the crop and prop exp fields, looks like it needs tidying:

summary(dat$CROPDMGEXP)
            ?      0      2      B      k      K      m      M 
618413      7     19      1      9     21 281832      1   1994 

summary(dat$PROPDMGEXP)
            -      ?      +      0      1      2      3      4      5      6      7      8      B      h      H 
465934      1      8      5    216     25     13      4      4     28      4      5      1     40      1      6 
     K      m      M 
424665      7  11330

A lot of missing ones, change the read to include na.strings=c(""," ","NA"))

summary(dat$PROPDMGEXP)
     -      ?      +      0      1      2      3      4      5      6      7      8      B 
     1      8      5    216     25     13      4      4     28      4      5      1     40 
     h      H      K      m      M   NA's 
     1      6 424665      7  11330 465934 

Next to tidy the data


```{r Cost tidy Damage Cost}
library(dplyr)

# Replace missing values with k

# dat2 <- data.frame(Event= NULL, PROP = NULL, CROP = NULL)
dat2 <- dat %>% select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
levels(dat2$PROPDMGEXP)
dat2$PROPDMGEXP[is.na] <- "K"
summary(dat2$PROPDMGEXP)
str(dat2)
# Change NA in a factor

# Set levels
levels(dat2$PROPDMGEXP) <- c("K", "M", "B")
# Replace the values
for (i in 1:length(DMG)){
        searchstring <- DMG[[i]]
        print(searchstring)
        dat2 <- dat %>%
                select(EVTYPE, PROPDMGEXP, CROPDMGEXP) %>%
                mutate(PROPDMGEXPtidy = as.character(PROPDMGEXP))
}

dat2 <- dat %>% select(EVTYPE, PROPDMGEXP, CROPDMGEXP, PROPDMG, CROPDMG, FATALITIES, INJURIES) %>%
        mutate(ID = seq_along(EVTYPE))

# Substitute H
DMG <- c("H", "h")
datHprop <- dat2 %>%
        filter(PROPDMGEXP %in% DMG) %>%
               mutate(PROPDMGEXPtidy = "H", DMGUOM = 100, PropCost = PROPDMG*100)
table(datHprop$PROPDMGEXPtidy)

# Substitute K
DMG <- c("K", "k","-", "?", "+", "0", "1", "2", "3", "4", "5")
datKprop <- dat2 %>%
        filter(PROPDMGEXP %in% DMG|is.na(PROPDMGEXP)) %>%
               mutate(PROPDMGEXPtidy = "K", DMGUOM = 1000, PropCost = PROPDMG*1000)
table(datKprop$PROPDMGEXPtidy)

datKcrop <- dat2 %>%
        filter(CROPDMGEXP %in% DMG|is.na(CROPDMGEXP)) %>%
               mutate(CROPDMGEXPtidy = "K", DMGUOM = 1000, CropCost = CROPDMG*1000)
table(datKcrop$CROPDMGEXPtidy)

# Substitute M
DMG <- c("M", "m", "6", "7", "8")
datMprop <- dat2 %>%
        filter(PROPDMGEXP %in% DMG) %>%
               mutate(PROPDMGEXPtidy = "M", DMGUOM = 1000000, PropCost = PROPDMG*1000000)
table(datMprop$PROPDMGEXPtidy)

datMcrop <- dat2 %>%
        filter(CROPDMGEXP %in% DMG) %>%
               mutate(CROPDMGEXPtidy = "M", DMGUOM = 1000000, CropCost = CROPDMG*1000000)
table(datMcrop$CROPDMGEXPtidy)

# Substitute B
DMG <- c("B", "b")
datBprop <- dat2 %>%
        filter(PROPDMGEXP %in% DMG) %>%
               mutate(PROPDMGEXPtidy = "B", DMGUOM = 1000000000, PropCost = PROPDMG*1000000000)
table(datBprop$PROPDMGEXPtidy)

datBcrop <- dat2 %>%
        filter(CROPDMGEXP %in% DMG) %>%
               mutate(CROPDMGEXPtidy = "B", DMGUOM = 1000000000, CropCost = CROPDMG*1000000000)
table(datBcrop$CROPDMGEXPtidy)
# Combine rows
datcrop <- rbind(datKcrop, datMcrop, datBcrop)
table(datcrop$CROPDMGEXPtidy)

datprop <- rbind(datHprop, datKprop, datMprop, datBprop)
table(datprop$PROPDMGEXPtidy)

# Combine Columns
# dat3 <- cbind(datprop, datcrop)
# rm(dat3)

dat3 <- merge(datprop, datcrop, by.x = "ID", by.y = "ID")

dat4 <- dat3 %>% mutate(ID, EVTYPE = EVTYPE.x, 
                        COST = PropCost+CropCost, 
                        FATALITIES = FATALITIES.x,
                        INJURIES = INJURIES.x) %>%
        select(ID, EVTYPE, COST, FATALITIES,INJURIES)

# Set factors

dat2 <- dat2 %>% mutate(PROPDMGEXPtidy = as.character(PROPDMGEXP))
table(dat2$PROPDMGEXPtidy)
dat2 <- dat2 %>% mutate(PROPDMGEXPtidy = 
                                grepl(searchstring, PROPDMGEXP, names(DMG[1]), 
                                     PROPDMGEXPtidy))

                select(EVTYPE, FATALITIES, INJURIES) %>%
                filter(grepl(searchstring, EVTYPE)) %>%
                mutate(Event = names(ET[i]))
# Dont do replace add new based on old

dat2$PROPDMGEXPtidy <- as.factor(dat2$PROPDMGEXPtidy)
rm(searchstring)
# Cost and multiple with the k,m,b values



# add total cost column with amount
dat3 <- dat %>%
        select(PROPDMGEXP, CROPDMGEXP)


```



