---
title: "Storm"
author: "K. Huisamen"
date: "31 May 2016"
output: html_document
---

#Synopsis
Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.


#Data Processing

```{r Load Data, cache=FALSE}
# Analysis start from the raw CSV file containing the data
# if exist then
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipfile <- "~/40 L&G/Coursera/ReproducibleResearch/Wk4_Project2/repdata-data-StormData.csv.bz2"
if (!exists("dat", envir = .GlobalEnv)){
        dat <- read.csv(zipfile)
}

head(dat)

# Tidy
# Convert EVTYPE all to upper
library(dplyr)
# library(tidyr)
dat <- dat %>%
        mutate(EVTYPE = toupper(EVTYPE))

Event.Types <- unique(dat$EVTYPE)
# Split event into
        #type (cause) - wind, fire, rain, mudslide
        #level - the degree of intensity, gale wind, etc
        #geolocation - sea, urban, mountain
# Explore


```



#Results
##Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r Most Harmful}

# Add level 1 type of event 
# wind

# water

# fire (incl heat)


# Quantify Health Risk
FatalityRisk <- 10
InjuryRisk <- 5
RiskLevels <- data.frame(Category = 5:1, Label = c("Disaster", "Significant", "Moderate", "Minor", "Minimal"))

HarmfulEvents <- dat %>%
        select(EVTYPE, FATALITIES, INJURIES) %>%
        group_by(EVTYPE) %>%
        summarise(Fatalities = sum(FATALITIES), Injuries = sum(INJURIES)) %>%
        as.data.frame %>%
        mutate(Risk = (FatalityRisk*Fatalities)+(InjuryRisk*Injuries)) %>%
        filter(Risk > 0) %>%
        arrange(desc(Risk)) %>%
        mutate(Category = ntile(Risk, 5)) #Split into 5 categories

HarmfulEvents <- merge(x = HarmfulEvents, y = RiskLevels, sort = FALSE)

# Add intensity
# Wind - Thunderstorm, High, Gusty

plot(HarmfulEvents$EVTYPE, HarmfulEvents$Fatalities)

```


##Across the United States, which types of events have the greatest economic consequences?

